---
title: "East River Conifer Species Mapping"
author: "Marshall Worsham"
date: "05/17/2024"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
    toc_depth: 2
---

```{r setup, include=F}
# Knitr options
knitr::opts_chunk$set(echo = F, warning = F, 
                      message = F, fig.fullwidth = TRUE, fig.align = 'center')

# Load config
config <- config::get(file=file.path('..', '..', '..', 'config',
                                     'config.yml'))
# Load packages
devtools::load_all()
load.pkgs(config$pkgs)

# Configure Google Drive and Maps auth
drive_auth(path=config$drivesa)
register_google(readLines(config$mapkey))
```

This notebook reports on several approaches to selecting tree stems for individual-scale species classification, uniting Kueppers, Worsham et al. forest inventory data with Falco et al. species classification map derived from the 2018 NEON spectroscopy mission. The approaches range from least to most conservative, that is, from preserving all trees to removing all but those in the uppermost canopy. Maps, agreement statistics, and agreement figures are shown for each approach.

## Pipeline
1. Align geolocated tree stem objects to classification raster.
2.
  - Define a buffer of radius *r* around each tree stem object, approximating crown area.
  - Segment trees using marker-controlled watershed segmentation algorithm to determine crown area.
3. Filter crown objects according to one of 4 specified procedures.
4. Plot a map of crown objects and classification data at an example site.
5. For each crown object, extract intersecting raster values by majority vote.
6. Compute accuracy statistics.
7. Generate agreement figures.

---

## Approaches
1. Height-dependent buffer and no filtering
2. Height-dependent buffer and filter on 90th percentile height or intersecting crowns < 90th percentile height
3. Height-dependent buffer and filter on 90th percentile height or non-intersecting crowns
4. Height-dependent buffer and filter on 90th percentile height
5. Static 3m buffer and filter on 90th percentile height
6. Segment tree-crown polygons and no filtering
7. Segment tree polygons and filter on 90th percentile height

---

## Ingest data
``` {r ingest, include=F}
# Ingest plot boundaries
plotsf <- load.plot.sf(path=as_id(config$extdata$plotid),
                       pattern=config$extdata$plotpattern)

# Ingest field data
tmpfile <- drive_download(
  as_id(config$extdata$invid),
  type='csv',
  path=file.path(tempdir(), config$extdata$invid),
  overwrite=T)$local_path

inv <- read.csv(tmpfile)

# Ingest Nicola's classification map
tmpfile <- drive_download(
  as_id(config$extdata$falcoid),
  path=file.path(tempdir(), config$extdata$falcoid),
  overwrite = T)$local_path

sp.class <- rast(tmpfile)
sp.class <- as.numeric(sp.class)
names(sp.class) <- 'Sp_Code'

# Ingest Nicola's species codes
tmpfile <- drive_download(
  as_id(config$extdata$classid),
  path=file.path(tempdir(), config$extdata$classid),
  overwrite=T)$local_path

sp.codes <- read.csv(tmpfile)
sp.codes <- sp.codes %>%
  mutate(Pixel_Code=as.numeric(Pixel_Code)) %>%
  dplyr::select(-c(Genus:Common))

# Ingest match data
modtrees <- read.csv(drive_download(
  as_id('1LCafch6gd2bYqYTT0eeW1o5QSHXv5O50'),
  path=file.path(tempdir(), 'opt_matches.csv'),
  overwrite = T)$local_path)

# Ingest las
infiles <- list.files(config$extdata$las_dec, full.names=T)
lascat <- readLAScatalog(infiles)

```

## Clean data
```{r clean, include=F}
# Subset field data to AOP sites
inv <- inv %>%
  filter(!Site_Name %in% c('XX-CAR1', 'XX-CAR2', 'XX-CAR3',
                           'XX-PLN1', 'XX-PLN2', 'SG-NWS1',
                           'XX-FAR1', 'ER-BME3'))

# Filter out field trees not meeting criteria
inv <- inv[grep('outside plot', inv$Comments, invert=T),] # Within plot bounds
inv <- inv[inv$Status == 'Live',] # Living stems
inv <- inv[!is.na(inv$Latitude | !is.na(inv$Longitude)),] # Have geolocation data

# Keep stem x,y,z data
stem.inv.xyz = data.frame('Tag_Number'=as.numeric(inv$Tag_Number),
                      'Height'=as.numeric(inv$Height_Avg_M),
                      'DBH'=as.numeric(inv$DBH_Avg_CM),
                      'Crown_Radius'=0.082*inv$Height_Avg_M + 0.5,
                      'Canopy_Position'=inv$Canopy_Position,
                      'X'=as.numeric(inv$Longitude),
                      'Y'=as.numeric(inv$Latitude),
                      'Sp_Code'=as.factor(inv$Sp_Code),
                      'Site_Name'=inv$Site_Name,
                      'Site_Tag'=paste0(inv$Site_Name, inv$Tag_Number))
stem.inv.xyz = stem.inv.xyz[!is.na(stem.inv.xyz$Height),]

# Turn stem.xyz into sf object
stem.sf.inv <- st_as_sf(stem.inv.xyz, coords=c('X', 'Y'), crs='EPSG:4326')
stem.sf.inv <- st_transform(stem.sf.inv, crs=st_crs(plotsf))

# FILTER OUT SUBORDINATE TREES
# Buffer around all trees
# Exclude if:
# A. a tree is < 80th pctl in height
# B. buffer intersects the buffer of another tree(s)
# C. any of those trees are larger

# Apply diameter-weighted buffer around all trees
stem.inv.buff <- st_buffer(stem.sf.inv, dist=stem.sf.inv$Crown_Radius,
                       endCapStyle = 'SQUARE', joinStyle='MITRE')

# Apply 3px buffer around all trees
stem.inv.buff.3m <- st_buffer(stem.sf.inv, dist=1, endCapStyle='SQUARE', joinStyle='MITRE')

# Create sparse matrix describing overlapping trees
stem.inv.overlap <- st_overlaps(stem.inv.buff)
stem.inv.within <- st_within(stem.inv.buff)

# Process modeled trees
modtrees <- modtrees %>%
  filter(src==1) %>%
  mutate(Site_Tag=paste0(site, obs)) %>%
  st_as_sf(coords=c('Xpred', 'Ypred'), crs='EPSG:32613') %>%
  left_join(stem.inv.xyz, by='Site_Tag') %>%
  mutate(Crown_Radius_Mod = 0.082*Zpred + 0.5)

st_crs(modtrees) <- st_crs(plotsf)

stem.sf <- modtrees

# Apply diameter-weighted buffer around all trees
stem.buff <- st_buffer(stem.sf, dist=stem.sf$Crown_Radius_Mod,
                       endCapStyle = 'SQUARE', joinStyle='MITRE')

# Apply 3px buffer around all trees
stem.buff.3m <- st_buffer(stem.sf, dist=3, endCapStyle='SQUARE', joinStyle='MITRE')

# Create sparse matrix describing overlapping trees
stem.overlap <- st_overlaps(stem.buff)
stem.within <- st_within(stem.buff)
```

## Make CHM
```{r}
# Make CHM

# Clip las to plot boundaries
lasplots <- clip_roi(lascat, plotsf)
lasplots <- lasplots[lapply(lasplots, nrow)>0]

# Rasterize canooy
chm.pitfree.05 <- lapply(lasplots, rasterize_canopy, 0.25, pitfree(), pkg = "terra")

# Smooth canopy
kernel <- matrix(1,5,5)
chm.smooth <- lapply(chm.pitfree.05, terra::focal, w = kernel, fun = mean, na.rm = TRUE)
```

## Prep dataset for plotting
```{r}
## Prep for plotting: subset and reclassify raster
gt1.sp <- crop(sp.class, ext(plotsf[plotsf$PLOT_ID=='ER-GT1',]))

# Reclassify raster using species codes
gt1.sp.fct <- reclass(gt1.sp, sp.codes)

# Pull basemap from google
gt1.bmap <- get_map(location=c(lon = -107.00595, lat = 38.9772),
                    zoom=20,
                    maptype = 'satellite',
                    source = 'google')

# Specify plotting palette
sp.pal <- PNWColors::pnw_palette('Sunset2', n=5)
```

---

## 1. Height dependent buffer and no filtering
The least conservative approach. Keep every LiDAR-detected tree. Set a height-dependent buffer around each object and attempt to extract species within the buffer using majority vote.

### Map
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.buff)
```

### Confusion matrix
```{r}
stems.comp1 <- get.spp(sp.class, stem.buff, sp.codes)
cm1 <- confusionMatrix(stems.comp1$Reference, stems.comp1$Classified)
cm1.overall <- cm1$overall

data.frame(cm1$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm1.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm1$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()
```

```{r}
pd1 <- plt.density(stems.comp1)
pb1 <- plt.bar(stems.comp1)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd1
```

### Agreement: frequency by species and size bin
```{r out.width='100%'}
pb1
```

---

## 2. Height-dependent buffer and filter on 90th percentile height or intersecting crowns < 90th percentile height

Keep any tree whose height is in the 90th percentile or higher. For all other trees, only keep them if (a) they don't intersect another tree's crown area or (b) if the tree whose area they intersect is not in the 90th percentile height.

1. Apply a height-dependent buffer of radius $r=0.03H + 0.5$, around every tree.
2. For each tree $t_i$:
  - if: $H \ge H_{90pctl}$, keep $t_i$.
  - else:
  - Test whether crown area sits entirely within another tree crown area.
      - if True: discard $t_i$
      - if False:
        - Test whether crown area of $t_i$ intersects any other tree crown area.
          - if True: test whether any of the trees $t_i$ intersects has $H \ge H_{90pctl}$
            - if True: discard $t_i$
            - if False: keep $t_i$
          - if False: keep $t_i$

```{r echo=T}
# Create canopy filter
canopy.filter.a <- unlist(lapply(1:nrow(stem.buff), \(i) {
  if(stem.buff[i,]$Zpred < quantile(stem.buff$Zpred, 0.9)) {
    if(length(stem.within[[i]])<=1) {
      tst <- stem.buff[i,]$Zpred > 0.9 * stem.buff[stem.overlap[[i]],]$Zpred &
        !any(stem.buff[stem.overlap[[i]],]$Zpred >= quantile(stem.buff$Zpred, 0.9))
      tst <- prod(tst)
    } else {
      tst <- 0 }
  } else {
    tst <- 1
  }
  as.logical(tst)
}
))

# Apply canopy filter to buffered stems
stem.filt.a <- stem.buff[canopy.filter.a,]
```

### Map
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.filt.a)
```

### Confusion matrix
```{r}
stems.comp2 <- get.spp(sp.class, stem.filt.a, sp.codes)
cm2 <- confusionMatrix(stems.comp2$Reference, stems.comp2$Classified)
cm2.overall <- cm2$overall

data.frame(cm2$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm2.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm2$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()

```

```{r}
pd2 <- plt.density(stems.comp2)
pb2 <- plt.bar(stems.comp2)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd2
```

### Agreement: frequency by species and size bin
```{r out.width='100%'}
pb2
```

---

## 3. Height-dependent buffer and filter on 90th percentile height or non-intersecting crowns
Keep any tree whose height is in the 90th percentile or higher. For all other trees, only keep them if (a) they don't intersect another tree's crown area.

1. Apply a height-dependent buffer of radius $r=0.042H + 0.675$, around every tree.
2. For each tree $t_i$:
  - if: $H \ge H_{90pctl}$, keep $t_i$.
  - else:
    - Test whether crown area of $t_i$ intersects any other tree crown area.
      - if True: discard $t_i$
      - if False: keep $t_i$

```{r echo=T}
# Create canopy filter
canopy.filter.b <- unlist(lapply(1:nrow(stem.buff), \(i) {
  if(stem.buff[i,]$Zpred < quantile(stem.buff$Zpred, .90)) {
    if(length(stem.within[[i]])<=1) {
      if(length(stem.overlap[[i]])) {
      tst <- 0
      } else {
        tst <- 1
      }
    } else {
      tst <- 0
    }
  } else {
  tst <- 1
  }
  as.logical(tst)
  }
))

# Apply canopy filter
stem.filt.b <- stem.buff[canopy.filter.b,]
```

### Map
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.filt.b)
```

### Confusion matrix
```{r}
stems.comp3 <- get.spp(sp.class, stem.filt.b, sp.codes)
cm3 <- confusionMatrix(stems.comp3$Reference, stems.comp3$Classified)
cm3.overall <- cm3$overall

data.frame(cm3$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm3.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm3$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()
```

```{r}
pd3 <- plt.density(stems.comp3)
pb3 <- plt.bar(stems.comp3)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd3
```

### Agreement: frequency by species and size bin
```{r out.width='100%'}
pb3
```

---

## 4. Height-dependent buffer and filter on 90th percentile height
Use a height-dependent buffer and keep only trees in the 90th percentile of height or higher.
```{r}
stem.filt.c <- stem.buff %>%
  group_by(Site_Name) %>%
  filter(Zpred >= quantile(Zpred, 0.9)) %>%
  ungroup() %>%
  mutate(Crown_Area=st_area(.))
```

### Map
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.filt.c)
```

### Confusion matrix
```{r}
stems.comp4 <- get.spp(sp.class, stem.filt.c, sp.codes)
cm4 <- confusionMatrix(stems.comp4$Reference, stems.comp4$Classified)
cm4.overall <- cm4$overall

data.frame(cm4$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm4.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm4$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()
```

```{r out.width='100%'}
pd4 <- plt.density(stems.comp4)
pb4 <- plt.bar(stems.comp4)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd4
```

### Agreement: frequency by species and size bin
```{r}
pb4
```

---

## 5. Static 3m buffer and filter on 90th percentile height
Apply a static 3px buffer around tree objects and keep only trees in the 90th percentile of height or higher.
```{r}
canopy.filter.d <- unlist(lapply(1:nrow(stem.buff), \(i) {

  if(stem.buff[i,]$Zpred < quantile(stem.buff$Zpred, .90)) {
  tst <- 0
  } else {
    tst <- 1
  }
  as.logical(tst)
}))

stem.filt.d <- stem.buff.3m[canopy.filter.d,]
```

### Map
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.filt.d)
```

### Confusion matrix
```{r}
stems.comp5 <- get.spp(sp.class, stem.filt.d, sp.codes)
summary(stems.comp5$Classified)
cm5 <- confusionMatrix(stems.comp5$Reference, stems.comp5$Classified)
cm5.overall <- cm5$overall

data.frame(cm5$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm5.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm5$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()

```

```{r}
pd5 <- plt.density(stems.comp5)
pb5 <- plt.bar(stems.comp5)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd5
```

### Agreement: frequency by species and size bin
```{r out.width='100%'}
pb5
```

---

## 6. Segment tree-crown polygons and no filtering
Segment crowns and keep all of them.
```{r}
# Get tree crown objects at plots
stem.sf$treeID <- 1:nrow(stem.sf)
crowns_ras <- lapply(chm.smooth, mcws, treetops = stem.sf, minHeight = 1.3)

# Plot CHM
crowns_poly <- lapply(chm.smooth, mcws, treetops = stem.sf, minHeight = 3, format='polygons')

stem.seg <- do.call('rbind', crowns_poly)
stem.seg <- st_join(st_as_sf(stem.seg), stem.sf, by=c('treeID'))

# Estimate crown area
stem.seg <- stem.seg %>%
  mutate(Crown_Area=st_area(.))

# Shrink crowns
stem.seg <- st_buffer(stem.seg, -0.1*sqrt(stem.seg$Crown_Area))
```

### Maps

#### CHM with all segmented crowns
```{r}
plot(chm.smooth[[10]], xlab = "", ylab = "", xaxt='n', yaxt = 'n')
plot(crowns_poly[[10]]$geometry, border = "blue", lwd = 0.5, add = TRUE)
```

#### Species with all segmented crowns
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.seg)
```

### Confusion matrix
```{r}
stems.comp6 <- get.spp(sp.class, stem.seg, sp.codes)
cm6 <- confusionMatrix(stems.comp6$Reference, stems.comp6$Classified)
cm6.overall <- cm6$overall

data.frame(cm6$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm6.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm6$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()
```

```{r out.width='100%'}
pd6 <- plt.density(stems.comp6)
pb6 <- plt.bar(stems.comp6)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd6
```

### Agreement: frequency by species and size bin
```{r}
pb6
```

---

## 7. Segment tree-crown polygons and filter on 90th percentile height
Segment crowns but use only trees in the 90th percentile of height or higher.
```{r}
# Get tree crown objects at plots
stem.sf$treeID <- 1:nrow(stem.sf)
crowns_ras <- lapply(chm.smooth, mcws, treetops = stem.sf, minHeight = 1.3)

# Plot CHM
crowns_poly <- lapply(chm.smooth, mcws, treetops = stem.sf, minHeight = 3, format='polygons')

stem.seg <- do.call('rbind', crowns_poly)
stem.seg <- st_join(st_as_sf(stem.seg), stem.sf, by=c('treeID'))

# Filter only top trees
stem.filt.e <- stem.seg %>%
  group_by(Site_Name) %>%
  filter(Zpred >= quantile(Zpred, 0.9)) %>%
  ungroup() %>%
  mutate(Crown_Area=st_area(.))

# Shrink crowns
stem.filt.e <- st_buffer(stem.filt.e, -0.1*sqrt(stem.filt.e$Crown_Area))
```

### Maps

#### CHM with all segmented crowns
```{r}
plot(chm.smooth[[10]], xlab = "", ylab = "", xaxt='n', yaxt = 'n')
plot(crowns_poly[[10]]$geometry, border = "blue", lwd = 0.5, add = TRUE)
```

#### Species with segmented crowns post-height filtering
```{r out.width='100%'}
mapit(gt1.sp.fct, stem.filt.e)
```

### Confusion matrix
```{r}
stems.comp7 <- get.spp(sp.class, stem.filt.e, sp.codes)
cm7 <- confusionMatrix(stems.comp7$Reference, stems.comp7$Classified)
cm7.overall <- cm7$overall

data.frame(cm7$table) %>%
  pivot_wider(names_from=Reference, values_from=Freq) %>%
  mutate(grp='Reference') %>%
  gt(rowname_col='Prediction',
     groupname_col='grp') %>%
  rm_stubhead() %>%
  tab_spanner(label='Classification',
              columns=ABLA:UNKN) %>%
  tab_options(row_group.as_column = TRUE)

data.frame(t(round(cm7.overall,2)))[c(2,1,3,4)] %>%
  flextable()

data.frame(round(cm7$byClass,2)) %>%
  rownames_to_column(var='Class') %>%
  flextable()
```

```{r out.width='100%'}
pd7 <- plt.density(stems.comp7)
pb7 <- plt.bar(stems.comp7)
```

### Agreement: kernel density by species and height
```{r out.width='100%'}
pd7
```

### Agreement: frequency by species and size bin
```{r}
pb7
```

### Write Approach 7 data table for reproducible tables
```{r}
stems.comp7.out <- stems.comp7 %>%
  st_set_geometry(NULL)

write.csv(stems.comp7.out, file.path(config$data$int, 'spp_class_data.csv'), row.names=F)
```


---

## Compare accuracy

```{r}
acc.df <- bind_rows(cm1.overall, cm2.overall, cm3.overall,
                cm4.overall, cm5.overall, cm6.overall, cm7.overall) %>%
  mutate(across(everything(), ~round(.,2)),
         Run = c('Ht-Dep Buffer - Least Conserv', 'Ht-Dep Buffer - Less Conserv', 'Ht-Dep Buffer - More Conserv',
                 'Ht-Dep Buffer - Most Conserv', 'Static Buffer - Most Conserv',
                 'Segmented Trees - Least Conserv', 'Segmented Trees - Most Conserv')) %>%
  dplyr::select(8, 2, 1, 3, 4)

acc.df %>%
  flextable()
```
