---
title: "Add stem geolocations to inventory datasheets"
author: "Marshall Worsham - UC Berkeley - worsham@berkeley.edu"
date: "2/28/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Contents

1 - [Front matter](#front)<br>
2 - [Workspace configuration](#workspace)<br>
3 - [Ingest data](#ingest)<br>
4 - [Clean data](#clean)<br>
5 - [](#)<br>
6 - [](#)<br>
7 - [](#)<br>

---
## Front matter <a id='front'></a>

Reads in tree stem locations from shapefile and appends longitude-latitude values in CRS EPSG:4326 to forest inventory datasheets.

## Workspace configuration <a id='workspace'></a>
```{r workspace,, include=F}
# Load config
config <- config::get(file=file.path('config', 'config.yml'))

# Load local helper functions and packages
devtools::load_all()
load.pkgs(config$pkgs)

# Configure drive oauth
drive_auth_configure(path=config$driveauth, api_key=readLines(config$drivekey))
drive_auth(email=config$email)
```

## Ingest data <a id='ingest'></a>
Import inventory data from Google Drive.
```{r ingestinv}
# Find on Drive
invsheets <- drive_find(pattern=config$extdata$invpattern, type='spreadsheet')
invnames <- invsheets$name
invids <- invsheets$id

# Download data to tmp directory and read to list using local function
inventory <- lapply(invnames, load.inventory)
```

Import tree geolocation shapefiles from Google Drive.
```{r ingesttrees}
# Read from Drive using local function
# trees <- load.trees(
#   path=as_id(config$extdata$stemid),
#   pattern=config$extdata$stempattern)

# Read from project's processed data directory
trees <- lapply(
  list.files(
    config$dat_pro,
    pattern='shp',
    full.names=T
  ),
  munge.trees
)
```

TODO: SEE IF THIS IS STILL NECESSARY BASED ON CHANGES IN 02.
Import raw tree geolocation GPS data from Google Drive. We use this for associating a filename with geolocation records. NOTE: Because this is a large request, it sometimes returns a cryptic https error. I think it has to do with request rate limits on a shared GCS credential that tidyverse/googledrive run in the background. Keep trying till it works.

```{r ingestgps}
# rgps <- load.rawgps(
#   as_id(config$extdata$rawgpsid),
#   pattern='Project',
#   nmax=950)
```

## Clean data <a id='clean'></a>

```{r}
# Row bind inventory data
inv.df <- bind_rows(inventory)

# Row bind tree geolocation data
trees.df <- bind_rows(trees)

# Row bind raw gps data
rawgps <- bind_rows(rawgps)

# Reassign rownames sequentially
rownames(inv.df) <- NULL
rownames(trees.df) <- NULL
rownames(rawgps) <- NULL

# Create a concatenated site-tag column to make unique site-tree IDs
# (A few tag sequences occur in more than one plot)
inv.df$SiTag <- paste0(inv.df$Site_Name, '_', inv.df$Tag_Number)
trees.df$SiTag <- paste0(trees.df$Site_Name, '_', trees.df$Tag_Number)

```

Review.
```{r}
head(inv.df)
head(trees.df)
```

Filter to 2021 records only. 
```{r}
# Filter to 2021 records only
data21 <- inv.df %>%
  filter((Census_End > '2021-01-01') & (Census_End < '2021-12-31'))
```

Review.
```{r}
unique(data21$Census_Start)
unique(data21$Site_Name)
dim(data21)
```

Find duplicate records and drop those not needed.
```{r}
dupes <- trees.df[duplicated(trees.df$SiTag),]
trees.df[trees.df$Tag_Number %in% dupes$Tag_Number,]
```

## Merge inventory records with tree tag records

First, perform the merge on the column `SiTag`.
```{r}
# Perform the merge on `SiTag`
invtrees <- merge(data21, trees.df, by = 'SiTag', all = T)
```

Next, make sure each record is attached to the correct site, and that each has a uniquely valid tag number

```{r}
# Ensure each record is attached to correct site
invtrees[is.na(invtrees$Site_Name.x),]$Site_Name.x <- invtrees[
  is.na(invtrees$Site_Name.x),]$Site_Name.y
invtrees[is.na(invtrees$Site_Name.x),]$Site_Name.x

# Ensure uniquely valid tag numbers
invtrees[is.na(invtrees$Tag_Number.x),]$Tag_Number.x <- invtrees[is.na(invtrees$Tag_Number.x),]$Tag_Number.y
```

Check if any tag numbers are NA
```{r}
dim(invtrees[is.na(invtrees$Tag_Number.x),]) # Should be 5
```

Review.
```{r}
# Find new duplicates in merged dataset
invtrees[duplicated(invtrees$SiTag),] # There shouldn't be any
tagdupes <- invtrees[duplicated(invtrees$Tag_Number.x),] 
dim(tagdupes) # There are several; it's OK

# Check how many observations are missing coordinates
length(which(is.na(invtrees$X))) # Should = 748

# Inspect
# View(invtrees[c('Site_Name.x', 'Tag_Number.x', 'Geotagged', 'Geotag_Association_Ref', 'Geotag_Association_Dist', 'Geotag_Association_Dir', 'X')])
```

Assign longitude, latitude, and GPS filename values from shapefiles.
```{r}
# Longitude
invtrees$Longitude <- invtrees$X

# Latitude
invtrees$Latitude <- invtrees$Y

# Filename
invtrees$GPS_File_Name <- invtrees$Filename
```

Drop unnecessary columns.
```{r}
invtrees <- select(
  invtrees,
  c(
    'Site_Name.x':'Geotag_Date',
    'Latitude',
    'Longitude',
    'GPS_File_Name':'Entry_Personnel',
  )
)
colnames(invtrees)
colnames(invtrees)[which(names(invtrees) %in% c(
    'Site_Name.x',
    'Tag_Number.x',
    'Sp_Code.x'))] <- c(
      'Site_Name', 
      'Tag_Number', 
      'Sp_Code')
```

```{r}
length(invtrees) == 44 #Check that there are 44 columns in resulting df
```

## Split merged records <a id='split'></a>

Split merged records into separate dataframes by site name, and store in list.
```{r}
sepdfs <- split(invtrees, with(invtrees, Site_Name), drop = F)
```

Look at how total numbers match up

```{r}
check.fin <- invtrees %>%
  group_by(Site_Name) %>%
  summarize(fin.n=n())

check.fin.u <- invtrees %>%
  group_by(Site_Name) %>%
  summarize(fin.u=n_distinct(Tag_Number))

check.org <- inv.df %>%
  group_by(Site_Name) %>%
  summarize(org.n=n())

check.org.u <- inv.df %>%
  group_by(Site_Name) %>%
  summarize(org.u=n_distinct(Tag_Number))

checks <- list(
  check.org,
  check.org.u,
  check.fin,
  check.fin.u)

check.n <- checks %>%
  reduce(
    inner_join,
    by='Site_Name'
  )

View(check.n)
```

## Final checks on each site's df <a id='finalcheck'></a>

### CC-CVN1 final check
```{r}
cc_cvn1 <- sepdfs$`CC-CVN1`
cc_cvn1[duplicated(cc_cvn1$Tag_Number),] # 1 duplicate
cc_cvn1[cc_cvn1$Tag_Number %in% cc_cvn1[duplicated(cc_cvn1$Tag_Number),'Tag_Number'],]
cc_cvn1 <- cc_cvn1[!duplicated(cc_cvn1$Tag_Number),]
nrow(cc_cvn1[(is.na(cc_cvn1$Longitude)) | (is.na(cc_cvn1$Latitude)), ]) # Missing 28 coordinates
nrow(cc_cvn1[((is.na(cc_cvn1$Longitude)) | (is.na(cc_cvn1$Latitude))) & (cc_cvn1$Geotagged==T),]) # 3 missing that we supposedly tagged
row.names(cc_cvn1) <- NULL
View(cc_cvn1)
```

### CC-CVN2 final check
```{r}
cc_cvn2 <- sepdfs$`CC-CVN2`
cc_cvn2 <- cc_cvn2[!is.na(cc_cvn2$Tag_Number),]
cc_cvn2[duplicated(cc_cvn2$Tag_Number),] # Two duplicates (1786, 1987)
cc_cvn2 <- cc_cvn2[!duplicated(cc_cvn2$Tag_Number),]
cc_cvn2[cc_cvn2$Tag_Number %in% c(1786, 1987),]
nrow(cc_cvn2[(is.na(cc_cvn2$Longitude)) | (is.na(cc_cvn2$Latitude)), ]) # Missing 16 coordinates
nrow(cc_cvn2[((is.na(cc_cvn2$Longitude)) | (is.na(cc_cvn2$Latitude))) & (cc_cvn2$Geotagged==T),]) # 10 missing that we supposedly tagged
row.names(cc_cvn2) <- NULL
```

### CC-CVS1 final check
```{r}
cc_cvs1 <- sepdfs$`CC-CVS1`
cc_cvs1[duplicated(cc_cvs1$Tag_Number),] # One duplicate (1196)
cc_cvs1[cc_cvs1$Tag_Number %in% c(1196),]
cc_cvs1 <- cc_cvs1[!duplicated(cc_cvs1$Tag_Number),]
nrow(cc_cvs1[(is.na(cc_cvs1$Longitude)) | (is.na(cc_cvs1$Latitude)), ]) # Missing 69 coordinates
nrow(cc_cvs1[((is.na(cc_cvs1$Longitude)) | (is.na(cc_cvs1$Latitude))) & (cc_cvs1$Geotagged==T),]) # 1 missing that we supposedly tagged
cc_cvs1 <- cc_cvs1[!is.na(cc_cvs1$Tag_Number),]
row.names(cc_cvs1) <- NULL
# View(cc_cvs1)
```

### CC-EMN1 final check
```{r}
cc_emn1 <- sepdfs$`CC-EMN1`
cc_emn1[duplicated(cc_emn1$Tag_Number),] # No duplicates
nrow(cc_emn1[(is.na(cc_emn1$Longitude)) | (is.na(cc_emn1$Latitude)), ]) # Missing 61 coordinates
nrow(cc_emn1[((is.na(cc_emn1$Longitude)) | (is.na(cc_emn1$Latitude))) & (cc_emn1$Geotagged==T),]) # 1 missing that we supposedly tagged
cc_emn1 <- cc_emn1[!is.na(cc_emn1$Tag_Number),]
row.names(cc_emn1) <- NULL
# View(cc_emn1)
```

### SG-NES1 final check
```{r}
sg_nes1 <- sepdfs$`SG-NES1`
# View(sg_nes1)
sg_nes1 <- sg_nes1[!is.na(sg_nes1$Tag_Number),]
sg_nes1[duplicated(sg_nes1$Tag_Number),] # 5 duplicates
sg_nes1[sg_nes1$Tag_Number %in% c(4456, 4589, 4628, 4774),]
sg_nes1 <- sg_nes1[!duplicated(sg_nes1$Tag_Number),]
sg_nes1[sg_nes1$Tag_Number %in% 4386,]$Latitude <- 38.93293972
sg_nes1[sg_nes1$Tag_Number %in% 4386,]$Longitude <- -106.98628092
nrow(sg_nes1[(is.na(sg_nes1$Longitude)) | (is.na(sg_nes1$Latitude)), ]) # Missing 219 coordinates
# View(sg_nes1[((is.na(sg_nes1$Longitude)) | (is.na(sg_nes1$Latitude))) & (sg_nes1$Geotagged==T),]) # 10 missing that we supposedly tagged
row.names(sg_nes1) <- NULL
# View(sg_nes1)
```

### SG-NES3 final check
```{r}
sg_nes3 <- sepdfs$`SG-NES3`
sg_nes3 <- sg_nes3[!is.na(sg_nes3$Tag_Number),]
sg_nes3[duplicated(sg_nes3$Tag_Number),] # 2 duplicates
sg_nes3[sg_nes3$Tag_Number %in% c(8149, 8151),]
sg_nes3 <- sg_nes3[!duplicated(sg_nes3$Tag_Number),]
nrow(sg_nes3[(is.na(sg_nes3$Longitude)) | (is.na(sg_nes3$Latitude)), ]) # Missing 167 coordinates
nrow(sg_nes3[((is.na(sg_nes3$Longitude)) | (is.na(sg_nes3$Latitude))) & (sg_nes3$Geotagged==T),]) # 10 missing that we supposedly tagged
row.names(sg_nes3) <- NULL
# View(sg_nes3)
```

### SR-PVG1 final check
```{r}
sr_pvg1 <- sepdfs$`SR-PVG1`
sr_pvg1 <- sr_pvg1[!is.na(sr_pvg1$Tag_Number),]
sr_pvg1[duplicated(sr_pvg1$Tag_Number),] # 0 duplicates
sr_pvg1 <- sr_pvg1[!duplicated(sr_pvg1$Tag_Number),]
nrow(sr_pvg1[(is.na(sr_pvg1$Longitude)) | (is.na(sr_pvg1$Latitude)), ]) # Missing 3 coordinates
nrow(sr_pvg1[((is.na(sr_pvg1$Longitude)) | (is.na(sr_pvg1$Latitude))) & (sr_pvg1$Geotagged==T),]) # none missing that we supposedly tagged
row.names(sr_pvg1) <- NULL
# View(sr_pvg1)
plot(sr_pvg1$Latitude, sr_pvg1$Longitude)
```

### XX-CAR1 final check
```{r}
xx_car1 <- sepdfs$`XX-CAR1`
xx_car1 <- xx_car1[!is.na(xx_car1$Tag_Number),]
xx_car1[duplicated(xx_car1$Tag_Number),] # 0 duplicates
xx_car1 <- xx_car1[!duplicated(xx_car1$Tag_Number),]
nrow(xx_car1[(is.na(xx_car1$Longitude)) | (is.na(xx_car1$Latitude)), ]) # Missing 96 coordinates
nrow(xx_car1[((is.na(xx_car1$Longitude)) | (is.na(xx_car1$Latitude))) & (xx_car1$Geotagged==T),]) # 11 missing that we supposedly tagged
row.names(xx_car1) <- NULL
# View(xx_car1)
plot(xx_car1$Latitude, xx_car1$Longitude)
```

### XX-PLN1 final check
```{r}
xx_pln1 <- sepdfs$`XX-PLN1`
xx_pln1[duplicated(xx_pln1$Tag_Number),] # 0 duplicates
# View(xx_pln1[xx_pln1$Tag_Number %in% xx_pln1[duplicated(xx_pln1$Tag_Number),'Tag_Number'],])
xx_pln1 <- xx_pln1[!duplicated(xx_pln1$Tag_Number),]
nrow(xx_pln1[(is.na(xx_pln1$Longitude)) | (is.na(xx_pln1$Latitude)), ]) # Missing 258 coordinates
nrow(xx_pln1[((is.na(xx_pln1$Longitude)) | (is.na(xx_pln1$Latitude))) & (xx_pln1$Geotagged==T),]) # 184 missing that we supposedly tagged
xx_pln1 <- xx_pln1[!is.na(xx_pln1$Tag_Number),]
row.names(xx_pln1) <- NULL
# View(xx_pln1)
plot(xx_pln1$Latitude, xx_pln1$Longitude)
```

### XX-PLN2 final check
```{r}
xx_pln2 <- sepdfs$`XX-PLN2`
xx_pln2[duplicated(xx_pln2$Tag_Number),] # 0 duplicates
# View(xx_pln2[xx_pln2$Tag_Number %in% xx_pln2[duplicated(xx_pln2$Tag_Number),'Tag_Number'],])
xx_pln2 <- xx_pln2[!duplicated(xx_pln2$Tag_Number),]
nrow(xx_pln2[(is.na(xx_pln2$Longitude)) | (is.na(xx_pln2$Latitude)), ]) # Missing 24 coordinates
nrow(xx_pln2[((is.na(xx_pln2$Longitude)) | (is.na(xx_pln2$Latitude))) & (xx_pln2$Geotagged==T),]) # 5 missing that we supposedly tagged
xx_pln2 <- xx_pln2[!is.na(xx_pln2$Tag_Number),]
row.names(xx_pln2) <- NULL
# View(xx_pln2)
plot(xx_pln2$Latitude, xx_pln2$Longitude)
# View(9iosepdfs$`ER-APL1`)
```
n 
## Write out<a id='write'></a>
```{r}

cleandfs <- list(cc_cvn1,
                 cc_cvn2,
                 cc_cvs1,
                 cc_emn1,
                 sg_nes1,
                 sg_nes3,
                 sr_pvg1,
                 xx_car1,
                 xx_car3,
                 xx_pln1,
                 xx_pln2)

for(d in cleandfs){
  write.csv(
    d,
    file.path(
      wkdir,
      'Forest_Inventory_Dataset',
      'Output',
      'CleanInvLatLon',
      paste0('clean_',
             d[1,1],
             '.csv')))
  print(d[1,1])
}
```

## Plot for one example
```{r}
ggplot(xx_pln2, aes(x=Longitude, y=Latitude, size=pi*(DBH_Avg_CM/2)^2)) +
  geom_point(color='darkgreen', fill='white', shape=21, alpha=0.3) +
  labs(size='Area')
```
